% PT
%\documentclass[phd,twoside,brazil]{ThesisPUC}
\documentclass[phd,twoside,british]{ThesisPUC_uk}

\usepackage{verbatim}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{alltt}

\usepackage[T1]{fontenc}
%\usepackage{lmodern}
%\usepackage[utf8]{inputenc}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{dsfont}

\usepackage{color}
\usepackage{xcolor,colortbl}
\definecolor{light}{gray}{0.90}
\definecolor{dark}{gray}{0.30}
%\definecolor{light}{rgb}{.90,.90,.90}
\definecolor{darkgreen}{rgb}{0,.50,0}
\definecolor{darkblue}{rgb}{0,0,.50}
\definecolor{darkred}{rgb}{.50,0,0}
\definecolor{darkpur}{rgb}{.50,0,.50}
\usepackage{listings}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\footnotesize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    %showspaces=false, % show spaces adding particular underscores
    %showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=t, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    emph={par,or,do,end,loop,await,emit,input,event,call,with,command,%
          var,and,not,then,else,native,return,pure,safe,nohold,finalize,%
          fin,mem,nop,PROC,PAR,CHAN,SIGNAL,async,sync,class},
    emphstyle={\bfseries},
    commentstyle=\color{dark},
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\Code}[1] {\texttt{#1}}

\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\nesc}{\emph{nesC}\xspace}
\newcommand{\FIN}{\code{finally}\xspace}
\newcommand{\DOFIN}{\code{do-finally}\xspace}

\newcommand{\footnoteremember}[2]{%
\footnote{#2}%
\newcounter{#1}%
\setcounter{#1}{\value{footnote}}%
}
\newcommand{\footnoterecall}[1]{%
\footnotemark[\value{#1}]%
}

\newcommand{\ST}{\1\xrightarrow[~n~]{}\1}
\newcommand{\BT}{\xRightarrow[(i,E)]{}}
\newcommand{\LL}{\langle}
\newcommand{\RR}{\rangle}
\newcommand{\DS}{\displaystyle}

\newcommand{\1}{\;}
\newcommand{\2}{\;\;}
\newcommand{\3}{\;\;\;}
\newcommand{\5}{\;\;\;\;\;}
\newcommand{\ten}{\5\5}
\newcommand{\twenty}{\ten\ten}

\definecolor{dark-red}{rgb}{0.2,0.15,0.15}
\definecolor{dark-blue}{rgb}{0.15,0.15,0.4}
\definecolor{medium-blue}{rgb}{0,0,0.5}
% PT
\hypersetup{
    colorlinks, linkcolor={dark-red},
    citecolor={dark-blue}, urlcolor={medium-blue}
}
%\usepackage[hidelinks]{hyperref}
%\hypersetup{hidelinks=true}
%\hypersetup{frenchlinks=true}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Francisco Figueiredo Goytacaz Sant'Anna}
\authorR{Sant'Anna, Francisco Figueiredo Goytacaz}
\adviser{Roberto Ierusalimschy}
\adviserR{Ierusalimschy, Roberto}
\coadviser{Noemi de La Roque Rodriguez}
\coadviserR{Rodriguez, Noemi de La Roque}

% PT
%\titulo{
%Concorr\^encia Segura em n\'ivel de Sistema para N\'os com Restri\c{c}\~oes de 
%Recursos com C\'eu
%}

\title{Safe System-level Concurrency on Resource-Constrained Nodes with C\'eu}

\day{12} \month{September} \year{2013}

\city{Rio de Janeiro}
\CDD{004}
\department{Inform\'atica}
\program{Computer Science}
\school{Centro T\'{e}cnico Cient\'{i}fico}
\university{Pontif\'{i}cia Universidade Cat\'{o}lica do Rio de Janeiro}
\uni{PUC--Rio}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\jury {
    %\jurymember{Edward Hermann Haeusler}
                %{Departamento de Inform\'{a}tica --- PUC-Rio}
    %\jurymember{Renato Fontoura de Gusm\~{a}o Cerqueira}
                %{Departamento de Inform\'{a}tica --- PUC-Rio}
    %\schoolhead{Jos\'{e} Eugenio Leal}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\resume {
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\acknowledgment{
I would like to thank my supervisors Roberto Ierusalimschy and Noemi Rodriguez 
for their time and knowledge sharing.

TTT: Olaf / Philippas

I'm also thankful to Luiz Fernando Soares, for my period in the Telem\'idia 
Lab, and all my teachers in CAp-UERJ, responsible for my education before 
college.

My studies were funded by CNPq and SAAB.
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% PT
\begin{comment}
\keywords {
    \key{Concorr\^encia}
    \key{Determinismo}
    \key{Sistemas Embarcados}
    \key{Esterel}
    \key{S\'incrono}
    \key{Reativo}
    \key{Redes de Sensores sem Fio}
}
\end{comment}

%\begin{comment}
\keywords{
  \key{Concurrency}%
  \key{Determinism}%
  \key{Embedded Systems}%
  \key{Esterel}%
  \key{Reactivity}%
  \key{Synchronous}%
  \key{Wireless Sensor Networks}%
}
%\end{comment}

% TODO muito corrido

%\begin{comment}
\abstract {
% TODO: split 2 sents.
Despite the continuous research to facilitate Wireless Sensor Networks 
development,
most safety analysis and mitigation efforts in concurrency are still left to 
developers, who must manage synchronization and shared memory explicitly.
% among activities explicitly.
%
We propose a system language that ensures safe concurrency by handling threats 
at compile time, rather than at runtime.
%
The synchronous and static foundation of our design allows for a simple 
reasoning about concurrency that enables compile-time analysis resulting in
deterministic and memory-safe programs.
%
As a trade-off, our design imposes limitations on the language expressiveness, 
such as doing computationally-intensive operations and meeting hard real-time 
responsiveness.
% defining, doing, executing, describing
%
To show that the achieved expressiveness and responsiveness is sufficient for a 
wide range of WSN applications, we implement widespread network protocols and 
the CC2420 radio driver.
%
The implementations show a reduction in source code size, with a penalty of 
memory increase below 10\% in comparison to \emph{nesC}.
%
Overall, we ensure safety properties for programs relying on high-level control 
abstractions that also lead to concise and readable code.
% in terms of total ROM/RAM.
%, and a negligible increase in CPU usage.
}
%\end{comment}

% PT

\begin{comment}
\resumo{
Apesar da pesquisa cont\'inua para facilitar a programa\c{c}\~ao de redes de 
sensores sem fio, a an\'alise de perigos de concorr\^encia ainda \'e de 
responsabilidade do programador, que deve tratar manualmente de quest\~oes como 
sincroniza\c{c}\~ao e mem\'oria compartilhada.
N\'os apresentamos uma linguagem de sistema que garante concorr\^encia segura 
tratando amea\c{c}as em tempo de compila\c{c}\~ao.
A fundamenta\c{c}\~ao est\'atica e s\'incrona da nossa abordagem permite um 
racioc\'inio mais simples sobre quest\~oes de concorr\^encia, permitindo uma 
an\'alise em tempo de compila\c{c}\~ao que garante programas determin\'isticos.
Como contra-partida, nosso modelo imp\~oe em termos da expressividade da 
linguagem, tais como para efetuar c\'alculos demorados, ou atender prazos 
estritos em tempo real.
N\'os implementamos diversos protocolos de rede conhecidos e o driver para o 
r\'adio CC2420 para mostrar que a expressividade e responsividade obtida com a 
linguagem \'e suficiente para uma gama consider\'avel de aplica\c{c}\~oes para 
redes de sensores.
As implementa\c{c}\~es mostram uma redu\c{c}\~ao de tamanho de c\'odigo, com um 
aumento de mem\'oria abaixo de 10\% em compara\c{c}\~ao com nesC.
O uso da linguagem proposta implica em diversas propriedades de seguran\c{c}a 
que se baseiam em abstra\c{c}\~oes de controle de alto n\'ivel, tamb\'em 
resultando em c\'odigo mais conciso e leg\'ivel.
}
\end{comment}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\tablesmode{fig} % none, fig, tab ou figtab
\setcounter{tocdepth}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

% PT
%\chapter{oi}

%\begin{comment}

\chapter{Introduction}
    \label{sec.intro}
    \input{intro.tex}

\chapter{Overview of programming models}
    \label{sec.models}
    \input{models.tex}

\chapter{The design of C\'eu}
    \label{sec.ceu}
    \input{ceu.tex}

\chapter{Demo applications}
    \label{sec.demos}
    \input{demo.tex}

\chapter{Evaluation}
    \label{sec.eval}
    \input{eval.tex}

\chapter{The semantics of C\'eu}
    \label{sec.formal}
    \input{formal.tex}

\chapter{The implementation of C\'eu}
    \label{sec.impl}
    \input{impl.tex}

%\chapter{Going Dynamic}
    %\section{Abstractions in C\'eu: Organisms}
    %\section{C\'eu organisms}
    %\label{sec.dyn}
    %\input{dyn.tex}

\chapter{Related work}
    \label{sec.related}
    \input{related.tex}

%\chapter{Ongoing work}
    %\label{sec.future}
    %\input{future.tex}

\chapter{Conclusion}
    \label{sec.conclusion}
    \input{conclusion.tex}

%\end{comment}

%\bibliographystyle{plain}
\bibliography{my,other}

%\appendix
%\chapter{The Full Semantics of C\'eu}
    %\label{app.formal}
    %\input{f.tex}

\end{document}
